---
title: "DNA methylation Illumina Mouse Array Quality Control Report"
author: "Emma Walker"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r setup, include=FALSE}

args<-commandArgs(trailingOnly = TRUE)
dataDir <- args[2]
#dataDir <- "/lustre/projects/Research_Project-191406/ARRand5xFAD/"
print(dataDir)

library(ENmix) # #https://www.bioconductor.org/packages/devel/bioc/vignettes/ENmix/inst/doc/ENmix.pdf 
library(ggplot2)
library(dplyr)
library(minfi)
library(gridExtra)
library(corrplot)


pheno <- paste0(dataDir, "0_metadata/sampleSheet.csv")
normDir <- paste0(dataDir, "2_normalised")
QCDir <- paste0(normDir, "/QC")  
configFile <- paste0(dataDir, "config.r") 

# load config.r
if (file.exists(configFile)) {
  source(configFile)
} else {
  stop("config.r file does not exist.")
}

# load sampleSheet
if (!file.exists(pheno)) {
  stop("Phenotype file not found at ", pheno)
}

sampleSheet <- read.csv(pheno, stringsAsFactors = F)

```

# Introduction

QC report for Mouse Data

Pipeline updated October 2024


# Loading data

An object of class ENmix::rgDataSet was made using the ENmix package and is loaded in alongside the sample sheet containing the pheno data for the mice, and QCmetrics object created by the calcMouseMethQCmetrics.r script


```{r, load QCmetrics,  warning=FALSE, message=FALSE, echo = FALSE}

print(QCDir)

load(file = paste0(QCDir, "/rgSet.rdat"))
load(file = paste0(QCDir, "/QCmetrics.rdat"))
load(file = paste0(QCDir, "/detP.rdat"))
load(file = paste0(QCDir, "/bsCon.rdat"))
load(file = paste0(QCDir, "/detP.rdat"))
load(file = paste0(QCDir, "/mraw.rdat"))

betas <- getB(mraw)

# define variables to plot
plotVars <- c(projVar, "Chip_ID")

# add column for neun and non neun celltypes
if(ctCheck){
  QCmetrics$NeuN <- ifelse(QCmetrics$Cell_Type == "NEUN", "NEUN", "OTHER")
}

QCmetrics$Chip_ID <- as.character(QCmetrics$Chip_ID)

# make sure age is numeric if it ia in the sample sheet
if("Age" %in% colnames(QCmetrics)){
  QCmetrics$Age <- as.numeric(QCmetrics$Age)
}

for(i in plotVars){
  if(is.character(QCmetrics[,i])){
    QCmetrics[,i] <- trimws(QCmetrics[,i])
  }
}

```

# Intensity check

The intensity check is often the biggest indicator of sample quality. The median methylated signal intensity and unmethylated signal intensity for each sample is calculated.

`r sum(!QCmetrics$IntensityPass)` samples fail at an intensity threshold of `r intensThresh`


# Intensity plots

```{r plotintensities, echo=FALSE, message=F}

# define function for catagorical datatype plots
factorIntensityPlots <- function(data, plotParam) {
  ggplot(data, aes(x = M.median, y = U.median, colour = as.factor(!!sym(plotParam)))) +
    geom_point() +
    xlab("Median M intensity") +
    ylab("Median U intensity") +
    ggtitle(plotParam)+
    labs(colour = plotParam)
}

# define function for continous datatype plots
continuousIntensityPlots <- function(data, plotParam) {
  ggplot(QCmetrics, aes(x = M.median, y = U.median, colour = !!sym(plotParam))) +
  geom_point() +
  xlab("Median M intensity") +
  ylab("Median U intensity") +
  scale_color_gradient(low = "blue", high = "red")+
  ggtitle(plotParam)+
  labs(colour = plotParam)
}


for(i in plotVars){
  if(!is.numeric(QCmetrics[,i])){
    print(factorIntensityPlots(QCmetrics, i))
  } else {
    print(continuousIntensityPlots(QCmetrics, i))
  }
}


```



```{r heatmaps, echo=FALSE, message=F}

QCmetrics$IntensityRatio <- QCmetrics$M.median/QCmetrics$U.median # previously M - U
QCmetrics$SentrixPosition <- factor(QCmetrics$Chip_Position)
QCmetrics$SentrixID <- factor(gsub("_.*", "", QCmetrics$Basename), levels=rev(unique(gsub("_.*", "", QCmetrics$Basename)))) #keeps the levels of the factor in current order rather than sorting numerically/alphabetically, also reverses this order as heatmaps plot bottom to top

#extract the legend (using a function found online)
g_legend <- function(a.gplot){
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  legend
}


plateHeatmap <- ggplot(data=QCmetrics, aes(x=SentrixPosition, y=SentrixID)) +
  scale_fill_gradientn(colours=colorRamps::matlab.like(100), limits=c(min(QCmetrics$U.median, QCmetrics$M.median),max(QCmetrics$U.median, QCmetrics$M.median))) +
  labs(x="", y="") +
  theme_minimal() + 
  coord_equal() +
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, hjust=1))

plot1 <- plateHeatmap +
  ggtitle("Median Methylated Intensity") +
  geom_tile(aes(fill=M.median), colour = "white") +
  theme(legend.position = "none")

plot2 <- plateHeatmap +
  ggtitle("Median Unmethylated Intensity") +
  geom_tile(aes(fill=U.median), colour = "white") +
  theme(legend.position = "none")

legendplot <- plateHeatmap + 
  geom_tile(aes(fill=U.median), colour = "white") +
  labs(fill="Intensity", alpha="Control") +
  scale_alpha_manual(values=c(1,1,1)) + 
  guides(alpha = guide_legend(override.aes = list(colour="black", pch=16)))

legend <- g_legend(legendplot)

chip_heatmaps <- grid.arrange(plot1, plot2, legend, ncol=3, widths=c(3/7, 3/7, 1/7))
ggsave("Methylation_intensity_CHIP.png", chip_heatmaps, device = "pdf")


```


# Bisulphite Conversion

```{r bisulphiteconversion, echo=FALSE}

ggplot(QCmetrics, aes(x = BsCon))+
  geom_histogram()+
  geom_vline(xintercept=bsConThresh, colour = "red", linetype="dashed")


```

`r sum(!QCmetrics$BsConPass)` samples fail at an intensity threshold of `r bsConThresh`


# P filter 

```{r, pfilt, echo=FALSE}

failedProbes <- rownames(detP)[((rowSums(detP > pFiltProbeThresh)/ncol(detP)) * 100) > pFiltSampleThresh]

```

`r length(failedProbes)` probes fail at a pvalue threshold of `r pFiltProbeThresh` in more than `r pFiltSampleThresh`% of samples





# Density Plots

```{r density plots, echo = FALSE, message = F}

betas <- betas[, QCmetrics$Basename]

densityPlot(betas, main = "All Betas")
densityPlot(betas, main = "Passed Samples", sampGroups = QCmetrics$PassQC1)

```



# Correlations between QC metrics

```{r, echo = FALSE, fig.width = 10, fig.height = 10}

colsToKeep <- c("M.median","U.median","BsCon", "pFiltPass", plotVars,
                  colnames(QCmetrics)[grep("PC", colnames(QCmetrics))])

colsToKeep <- colsToKeep[colsToKeep %in% colnames(QCmetrics)]

corDF <- QCmetrics[,colsToKeep]

# change to correct type

for(i in colsToKeep){
  if(!is.numeric(QCmetrics[,i])){
   corDF[,i] <- as.numeric(as.factor(corDF[,i]))
  } 
}



corrplot::corrplot(cor(corDF, use = "p"))

```


