# 10 Tissue Type Check

n.b this check is run on passed QC stage 1 samples only and is provided for information only. It does NOT exlude any further samples prior to normalisation.

```{r tCheckSetup, include=FALSE}
library(ENmix)
library(pheatmap)
library(pander)
library(RColorBrewer)
library(kableExtra)
library(data.table)

# load manifest

refDir <- args[3]
refDir <- "/lustre/home/ew367/mouseArray/"
manifest <- paste0(refDir, "MouseMethylation-12v1-0_A2.csv")

man <- fread(manifest, skip=7, fill=TRUE, data.table=F)

## filter samples
QCSum <- read.csv(file.path(QCDir, "passQCStatusStage1AllSamples.csv"), stringsAsFactors = F)
#QCSum <- na.omit(QCSum)
passQC <- QCSum$Basename[QCSum$PassQC1]
load(file = file.path(QCDir, "QCmetrics.rdat"))

nStart<-nrow(QCSum)
nExclude<-sum(!QCSum[,"PassQC1"])

#nCTStart<-table(QCmetrics$Tissue_Type)

QCmetrics<-QCmetrics[match(passQC, QCmetrics$Basename),]

#nCTMid<-table(QCmetrics$tissueType)

#tCounts<-cbind(nCTStart, nCTMid)
#colnames(tCounts)<-c("Profiled", "PassQCStage1")


# get betas, match to passed samples only and subset to just auto probes
load(file = file.path(QCDir, "mraw.rdat"))
betas <- getB(mraw)
rawbetas<-betas[,match(passQC, colnames(betas))]
auto.probes<-man$IlmnID[man$CHR != "X" & man$CHR != "Y" & man$CHR != "MT"]
rawbetas<-rawbetas[row.names(rawbetas) %in% auto.probes,]

# calculate betas.scores
pca <- prcomp(t(rawbetas))
betas.scores = pca$x
colnames(betas.scores) = paste(colnames(betas.scores), '_betas', sep='')
betas.pca<-pca$sdev^2/sum(pca$sdev^2)
betas.scores<-betas.scores[,which(betas.pca > 0.01)]



tissueTypes<-sort(unique(QCmetrics$Tissue_Type))
tissueCols<-brewer.pal(n = length(tissueTypes), name = "Paired")
tissueColsTrans<-rgb(t(col2rgb(tissueCols)), max = 255, alpha = 0.5)
#tissueCols<-c("darkgreen", "darkblue", "darkmagenta", "deeppink", "darkgray", "black", "yellowgreen", "orangered4")

## filter out NAs
rawbetas<-na.omit(rawbetas)

## filter out SNPs
if(length(grep("rs", rownames(rawbetas))) > 0){
  rawbetas<-rawbetas[-grep("rs", rownames(rawbetas)),]
}

sample_anno<-QCmetrics[,c("Age","Sex", "Tissue_Type")[c("Age","Sex", "Tissue_Type") %in% colnames(QCmetrics)]]
if("Age" %in% colnames(sample_anno)){
  sample_anno$Age<-as.numeric(sample_anno$Age)
}
rownames(sample_anno)<-QCmetrics$Basename

sample_anno<-sample_anno[,colSums(!is.na(sample_anno))> 0, drop=F]

sigma<-apply(rawbetas, 1, sd)



```

# Summary of Quality Control 

After performing quality control (QC) across all samples (n = `r nStart`), `r nExclude`  samples were excluded after stages 1 and 2 of the QC. 
The third stage of the QC pipeline will look to establish the success of the FACs sorting and check whether samples cluster by their labelled tissue type. 

```{r, echo = FALSE}

kbl(tCounts) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),font_size = 10)

```


# Quality Control: Clustering of Tissue Types

## Initial Visualisation 

First, we will produce a heatmap and hierarchical cluster across all samples that passed stages 1 and 2 of the QC to gauge how cleanly the tissue-types are clustering. The following heatmap is on based on 500 most variable autosomal probes (ranked by SD), we will force it to cut the hierarchical cluster into the same number of groups as tissue types. 

```{r hclustBefore, fig.width = 15, fig.height = 8, echo = FALSE}
## initial cluster
pheatmap(rawbetas[order(sigma, decreasing = TRUE)[1:500],], annotation_col = sample_anno,  show_colnames = FALSE, show_rownames = FALSE, cutree_cols = length(tissueTypes), main = "Pre Sample Type Filtering")


```


## Principal Component Analysis

We will use principal components to classify samples into tissue types and determine if the profile of the sample is representative of the sample it is labelled as. The figure below shows the PCA before any checking of tissue type.

```{r PCAbefore, fig.width = 15, fig.height = 8, echo = FALSE}
par(mfrow = c(1,2))
par(mar = c(4,4,0.5,0.5))
plot(betas.scores[,1], betas.scores[,2],col = tissueCols[as.factor(QCmetrics$Tissue_Type)], pch = 16, xlab = "PC1", ylab = "PC2", cex.axis = 1.5, cex.lab = 1.5)

plot(betas.scores[,3], betas.scores[,4],col = tissueCols[as.factor(QCmetrics$Tissue_Type)], pch = 16, xlab = "PC3", ylab = "PC4", cex.axis = 1.5, cex.lab = 1.5)

legend("bottomright", pch = 16, col = tissueCols, levels(as.factor(QCmetrics$Tissue_Type)))

```


We will also visualize which axes of variation separate the different tissue types. Note that at this stage, we haven't filtered out any incorrectly labelled samples so there may still be some noise.

```{r PCAboxplot,fig.width = 15, fig.height = 15, echo = FALSE}
par(mfrow = c(3,2))
par(mar = c(6,4,0.5,0.5))
par(mgp = c(2,0.75,0))
for(i in 1:ncol(betas.scores)){
  
  boxplot(betas.scores[,i] ~ QCmetrics$Tissue_Type, col = tissueCols, ylab = paste0("PC", i), xlab = "", cex.lab = 1.5, cex.axis = 1.5, xaxt = "n")
  text(x = 1:length(tissueTypes),
     y = par("usr")[3] - diff(par("usr")[3:4])*0.05,
     labels = tissueTypes,
     xpd = NA,
     ## Rotate the labels by 35 degrees.
     srt = 35,
     cex = 1.2, adj = 1)
  
}
```
